Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-22 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Revised Ch 8.1 through 8.5 per CF comments

The goal of this chapter is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study.  Computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop their own data or rely on loosely defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

 The FDS User Guide provides a method to determine how well a flow field will be resolved by using the non-dimensional expression ${D^*}/{\delta x}$, where $D^*$ is a characteristic fire diameter and $\delta x$ is the nominal size of a mesh or grid cell. This value has been referred to as the plume resolution index. Given that the fire plume region is the focus of this study, it is important to understand how changes in plume mesh resolution impact the results.

@@

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `thermocouple' results from FDS to see if the model of the thermocouple provides improved agreement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `thermocouple' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `thermocouple' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead may be generating the differences between the gas temperature and the measured temperature FDS values.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/FHNG01_RI_TC} \\
	\caption[Comparison of vertical centerline `thermocouple' temperatures above the natural gas burner with FDS simulated gas temperatures.]{Comparison of vertical centerline `thermocouple' temperatures above the natural gas burner with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20.}
	\label{FHNG01_RI_TC}
\end{figure}

@@

Reviewing the heat flux data documented in Chapter 2 from both the 0.95~m  and the 2.0~m from the vertical centerline of the burner positions provided average steady state heat flux values of approximately 2.0~kW/m$^2$ and 0.43~kW/m$^2$, respectively.  Applying the point source method, discussed in Chapter 7, conservative values for the radiative fraction were determined to be in the range of 0.25 to 0.28.  Values found in the literature for natural gas range from 0.15 to 0.25~\cite{Modak:1977,Hostikka:3}.   
Therefore it was of interest to conduct FDS simulations with radiative fractions ranging from 0.2 to 0.3.  This was based on the default value used in FDS and the upper bound of the literature values as this range covered the radiative fraction values estimated from the experiments.  The course grid, RI=5, was used for the FDS models.  The comparison of the results is shown in~\ref{FHNG01_HF_RI5}.  The radiative fraction of 0.22 provided the best agreement with the measurement.

@@

One of the outputs that can be gotten from FDS is a heat release rate per unit length (HRRPUL), it has been used to determine the height where energy stops being released along the vertical centerline of the gas burner~\cite{McDermott:2011}.  Estimates of the flame height were made by estimating the height where 0.95 to 0.99~$\%$ of the total energy had been released. The peak visible flame height range estimate was developed from the location where the flame exists only 10~\% of the time with the 95~\% certainty range.  The Table~\ref{tab:FDS_FH} displays estimated values from the models using all three RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, were not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range.

@@

One of the advantages of a computational fluid dynamics model such as FDS is the ability produce enough spatial data in 3 dimensions to allow the generation of visualizations.  The impact of the mesh resolution can be seen in the three images of the fire shown in Figure~\ref{FDS_Flame_Image_comp}.  The increased resolution leads to a flame structure that provides a better match to the look of the actual flame.  However in this scenario, the increased resolution within the flow field has not resulted in improved numerical agreement.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube with 2.4~m on a side. The models were run with two different sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were: thermal conductivity (0.28 W/m K), specific heat (1.00 kJ/kg K) and density (807 kg/m$^3$).  In addition FDS includes a default value of 0.9 for emissivity, this value should be representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density and as the paper is charred or burned there is a potential for a change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500~$^\circ$C~\cite{Gross:1985}.  While FDS can accommodate changes in thermo-conductivity and specific heat with respect to time, density and emissivity must be held constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throughout the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

As before, the FDS simulations included devices which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in locations similar to the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL will again be used as a means to gain insight into the peak flame height simulated by FDS.

@@

 There seem to be a couple of items that stick out from the graphs.  First, there appears to be a separation of the temperature data into three distinct sections.  The bottom three positions, 0.2~m through 0.6~m, represent the hottest section.  The next two measurement positions at 0.8~m and 1.0~m are similar in temperature.  The third section extends from 1.2 through 2.2~m above the burner and has the shape of a smooth curve decreasing in temperature with increasing elevation.  Comparing the temperatures to the flame height analysis in Chapter 2, the first section aligns with the median flame height, the second section aligns with the portion of the flame that is intermittent, 40~$\%$ of the time or less.  The upper most section of the temperature data aligns with the thermal plume above the flame.

@@

Reviewing the videos from a natural gas burner instrumented gypsum wallboard experiment with the burner positioned at 0.310~m from the wall, we can see examples of the flame movement.  In the top row of Figure~\ref{FDS_IWNG_Flame_comp} the three images exhibit points in time where the flame was moving toward the instrumented wall and away from the vertical array of thermocouples that are positioned above the center line of the burner.  The bottom row of images are from the same fire experiment looking at it from the front.  These images also show how the pulsations and movements of the flame change positions relative to the fixed positions of the thermocouples and heat flux gauges. 
Figure~\ref{FDS_IWNG_SMV_comp} has a front view and a side view Smokeview image from the FDS RI=10 simulation with the burner positioned at 0.31~m from the wall.  This simulation is comparable to the experiment used for Figure~\ref{FDS_IWNG_Flame_comp}.  While the simulation does not demonstrate the amount of eddy generation or side to side movement from the front, the image from the side does show that FDS is simulating that the flame is moving toward the wall, although not to the extent that the actual flame did.

@@

The flame height range was calculated using the HRRPUL from each of the FDS simulations at each position from the wall and for RI=5 and RI=10.  The peak visible flame height range estimate was developed from the location where the flame exists only 10~\% of the time with the 95~\% certainty range.  This information is presented in Table~\ref{tab:FDS_FH_IWNG} displays estimated values from the models using all three RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, were not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range at the position most remote from the wall.

@@

The comparisons were made for the natural gas burner fire positioned against a painted gypsum board wall.  Figure~\ref{FDS_FirePattern_comp} displays two Smokeview images.  In the image on the left the isothermal areas of the wall shown with a temperature range of 20~$^\circ$C to greater than 520~$^\circ$C.  The Smokeview image on the right has the `char' feature applied with the threshold temperature of 360~$^\circ$C to coincide with cone calorimeter ignition temperature data. The upper edge of the char area is approximately 1~m above the burner.  The upper edge of the 520~$^\circ$C area in the left image was approximately 0.8~m above the burner. This images were generated with the RI=10 model which has grid cells that were 0.4~m in size. The FDS images are compared with three photographs of fire patterns from the compartment fire pattern experiments documented in Chapter~6.  The three images from left to right exhibit a range of fire pattern heights from approximately 0.7 to 0.8~m.  The average height of the natural gas fire patterns from this open wall construction series was 0.74 with a 95~\% certainty range of $pm$~0.12~m. The simulated fire pattern is taller than the actual patterns.  However is the upper ignition temperature was used the height values between the actual and the simulated would have been similar.  The RI=10 simulation was used as a means to optimize the graphical resolution, even though the temperature range of the flames were shifted up relative to the RI=5 case.

>>>>

The goal of this chapter is to utilize a numerical tool to predict or simulate the results from some of the key experiments conducted in this study.  Computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge, more computer resources, and more input data to utilize correctly.  Currently modelers need to develop their own data or rely on loosely defined databases with the hope that the material tested in the database is similar to the specific material of interest.

@@

 The FDS User Guide provides a method to determine how well a flow field will be resolved by using the non-dimensional expression ${D^*}/{\delta x}$, where $D^*$ is a characteristic fire diameter and $\delta x$ is the nominal size of a mesh or grid cell. This value has been referred to as the plume resolution index~(RI). Given that the fire plume region is the focus of this study, it is important to understand how changes in plume mesh resolution impact the results.

@@

The comparisons above used the predicted gas temperature from the model. Another comparison can be made examining the `THERMOCOUPLE' results from FDS to see if the model of the thermocouple provides improved agreement with the temperatures measured via physical thermocouples.  This comparison is shown in Figure~\ref{FHNG01_RI_TC}.  Not surprisingly, the sub model of the thermocouple device shifted the FDS values slightly.  At temperatures greater than 700~$^\circ$C, the `THERMOCOUPLE' temperatures tended to be lower than the gas temperatures, perhaps due to radiant heat transfer losses. At temperatures less than 700~$^\circ$C, the `THERMOCOUPLE' temperatures tended to be a higher than the gas temperatures, perhaps due to gains from radiant heat transfer.  The best agreement along the vertical array of temperatures was provided by the RI=5 model.  Variables such as density and emissivity required for modeling the thermal response and heat transfer to the thermocouple bead may be generating the differences between the gas temperature and the measured temperature FDS values.

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/FHNG01_RI_TC} \\
	\caption[Comparison of vertical centerline `THERMOCOUPLE' temperatures above the natural gas burner with FDS simulated gas temperatures.]{Comparison of vertical centerline `THERMOCOUPLE' temperatures above the natural gas burner with FDS simulated gas temperatures generated with three different RIs 5, 10, and 20.}
	\label{FHNG01_RI_TC}
\end{figure}

@@

Reviewing the heat flux data documented in Chapter 2 from both the 0.95~m  and the 2.0~m from the vertical centerline of the burner positions provided average steady state heat flux values of approximately 2.0~kW/m$^2$ and 0.43~kW/m$^2$, respectively.  Applying the point source method, discussed in Chapter 7, values for the radiative fraction were determined to be in the range of 0.25 to 0.28.  Values found in the literature for natural gas range from 0.15 to 0.25~\cite{Modak:1977,Hostikka:3}.   
Therefore it was of interest to conduct FDS simulations with radiative fractions ranging from 0.2 to 0.3.  This was based on the default value used in FDS and the upper bound of the literature values as this range covered the radiative fraction values estimated from the experiments.  The course grid, RI=5, was used for the FDS models.  The comparison of the results is shown in Figure~\ref{FHNG01_HF_RI5}.  The radiative fraction of 0.22 provided the best agreement with the measurement.

@@

One of the outputs that can be gotten from FDS is a heat release rate per unit length (HRRPUL), it has been used to determine the height where energy stops being released along the vertical centerline of the gas burner~\cite{McDermott:2011}.  Estimates of the flame height were made by estimating the height where 95 to 99~$\%$ of the total energy had been released. To have a value, based on a measurement, for comparison to the HRRPUL the peak visible flame height range estimate was developed.  The peak visible flame height ranged from the location where the flame existed only 10~\% of the time with the 95~\% certainty range.  The Table~\ref{tab:FDS_FH} displays estimated values from the models using all three RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, were not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range.

@@

One of the advantages of a computational fluid dynamics model such as FDS is the ability produce enough spatial data in 3 dimensions to allow the generation of visualizations.  The impact of the mesh resolution can be seen in the three images of the fire shown in Figure~\ref{FDS_Flame_Image_comp}.  The increased resolution leads to a flame structure that provides a better match to the look of the actual flame.  However in this scenario, the increased resolution within the flow field has not resulted in improved numerical agreement with the measured values.

@@

The computational domain of the model was the same as the volume considered for the source fire simulations, a cube with 2.4~m on a side. The models were run with two different sized computational grid cells, an 80~mm (RI=5) grid and a 40~mm (RI=10) grid.  The face of the wall is located 1.95~m from the front of the domain.  
The wall was modeled with material properties for gypsum wallboard that were documented in Chapter 4.  The input properties were: thermal conductivity (0.15 W/m K), specific heat (1.00 kJ/kg K) and density (807 kg/m$^3$).  In addition FDS includes a default value of 0.9 for emissivity, this value should be representative of the painted surface of the gypsum wallboard. As noted earlier in this study, gypsum wallboard undergoes chemical and physical changes as it is heated.  These changes include the loss of homogeneity of the gypsum material, a change in density and as the paper is charred or burned there is a potential for a change in emissivity.  The thermal conductivity and the specific heat of the gypsum wallboard varies on the order of 20~$\%$ over the temperature range from 20 to 500~$^\circ$C~\cite{Gross:1985}.  While FDS can accommodate changes in thermo-conductivity and specific heat with respect to time, density and emissivity must be held constant~\cite{FDS_Users_Guide}.  For the purposes of this study, the input values will remain constant throughout the course of the simulations.  The only variable will be the distance between the edge of the burner and the wall.

As before, the FDS simulations included devices which are used to simulate thermocouples and total heat flux gauges.  These devices were positioned in the same locations as the locations used in the physical experiments, above the centerline of the burner and on the wall as shown in Chapter 4, Figure~\ref{Instrumented_GB_Wall}. In addition, HRRPUL will again be used as a means to gain insight into the peak flame height simulated by FDS.

@@

 There seem to be a couple of noticeable trends from the graphs.  First, there appears to be a separation of the temperature data into three distinct sections.  The bottom three positions, 0.2~m through 0.6~m, represent the hottest section.  The next two measurement positions at 0.8~m and 1.0~m are similar in temperature.  The third section extends from 1.2 through 2.2~m above the burner and has the shape of a smooth curve decreasing in temperature with increasing elevation.  Comparing the temperatures to the flame height analysis in Chapter 2, the first section aligns with the median flame height, the second section aligns with the portion of the flame that is intermittent, 40~$\%$ of the time or less.  The upper most section of the temperature data aligns with the thermal plume above the flame.

@@

Examples of the flame movement were found by reviewing the videos from a natural gas burner, instrumented gypsum wallboard experiment with the burner positioned at 0.310~m from the wall.  Figure~\ref{FDS_IWNG_Flame_comp} shows images from two different locations, from the front of the burner and from the right side of the burner.  In the top row of Figure~\ref{FDS_IWNG_Flame_comp} the three images exhibit points in time where the flame was moving toward the instrumented wall and away from the vertical array of thermocouples that are positioned above the center line of the burner.  The bottom row of images are from the same fire experiment looking at it from the front.  These images also show how the pulsations and movements of the flame change positions relative to the fixed positions of the thermocouples and heat flux gauges. 
Figure~\ref{FDS_IWNG_SMV_comp} has a front view and a side view Smokeview image from the FDS RI=10 simulation with the burner positioned at 0.31~m from the wall.  This simulation is comparable to the experiment used for Figure~\ref{FDS_IWNG_Flame_comp}.  While the simulation does not demonstrate the amount of eddy generation or side to side movement from the front, the image from the side does show that FDS is simulating that the flame is moving toward the wall, although not to the extent that the actual flame did.

@@

The flame height range was calculated using the HRRPUL from each of the FDS simulations at each position from the wall and for RI=5 and RI=10.  The peak visible flame height range estimate was developed from the location where the flame exists only 10~\% of the time with the 95~\% certainty range.  This information is presented in Table~\ref{tab:FDS_FH_IWNG} displays estimated values from the models using two of the RI values as well as the 95~$\%$ certainty range for the peak visible flame height from Chapter 2.  The course and medium grids, RI=5 and R=10, were not significantly different from the estimations of the peak visible flame height 95~$\%$ certainty value range at the position most remote from the wall.

@@

The comparisons were made for the natural gas burner fire positioned against a painted gypsum board wall.  Figure~\ref{FDS_FirePattern_comp} displays two Smokeview images.  In the image on the left the isothermal areas of the wall shown with a temperature range of 20~$^\circ$C to greater than 520~$^\circ$C.  The Smokeview image on the right has the `char' feature applied with the threshold temperature of 360~$^\circ$C to coincide with cone calorimeter ignition temperature data. The upper edge of the char area is approximately 1~m above the burner.  The upper edge of the 520~$^\circ$C area in the left image was approximately 0.8~m above the burner. This images were generated with the RI=10 model which has grid cells that were 0.04~m in size. The FDS images are compared with three photographs of fire patterns from the compartment fire pattern experiments documented in Chapter~6.  The three images from left to right exhibit a range of fire pattern heights from approximately 0.7 to 0.8~m.  The average height of the natural gas fire patterns from this open wall construction series was 0.74 with a 95~\% certainty range of $\pm$~0.12~m. The simulated fire pattern is taller than the actual patterns.  However if the upper ignition temperature was used the height values between the actual and the simulated would have been similar.  The RI=10 simulation was used as a means to optimize the graphical resolution, even though the temperature range of the flames were shifted up relative to the RI=5 case.
